To be FAIR: Theory Specification Needs an Update

### 2025-02-19 manuscript.Rmd
Add examples for relationship types (see #32)

We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
At present, the primary unit of social scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable *information artifacts* in their own right,
with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
Data sharing is a good example of a modular publishing practice that is  widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program.
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
which is used by many academic publishers.
Finally, the [DataCite Metadata Schema]() offers a standard way to document the nature of relationships between information artifacts.
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
and cross-reference the published paper with `relationType: IsSupplementTo`.
Similarly, FAIR theories can be connected to a specific paper which might serve as the theory's documentation and canonical reference.

>>>>

We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
At present, the primary unit of social scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable *information artifacts* in their own right,
with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program.
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
which is used by many academic publishers.
Finally, the [DataCite Metadata Schema]() offers a standard way to document the nature of relationships between information artifacts.
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
and cross-reference the published paper with `relationType: IsSupplementTo`.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`. Other types are useful for making relationships between multiple theory objects explicit: If an existing narrative theory is FAIRified without much own contribution, the FAIR theory would cross-reference the paper with the narrative theory as `relationType: IsDerivedFrom`; if substantial own contributions were necessary, `relationType:IsNewVersionOf` could be used. If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<!-- FS: I still think we should have a dedicated term for a digital FAIR theory object -->
